Full Waveform Inversion by Source Extension: Why it works

### 2020-03-28 repro/papers/dso1d/paper.tex
near final - need WRI ref

\maketitle
\begin{abstract}
An extremely simple example shows how extended modeling plus variable projection plus a suitable annihilator lead to constructive velocity updates.
\end{abstract}

\section{Introduction}
Full Waveform Inversion (FWI) is now well-established as
a useful tool for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10}, but its use is still
impeded by so-called ``cycle-skipping''. Because the computational
size of field inversion tasks is very large, only iterative local (descent)
minimization of the data misfit function is computationally
feasible. However such local descent methods tend to stagnate as
suboptimal and geologically uninformative earth models, unless initial
models are already quite close to optimal, in the sense of predicting
the arrival times of seismic events to within a small multiple of a
dominant wavelenght \cite[]{GauTarVir:86,Plessix:10}.

This paper concerns one of the many ideas that have been advanced to
overcome cycle-skipping, namely so-called extended inversion
\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
of freedom are provided to the modeling process, in the hope of
opening up more effective routes to geologically informative models
with acceptable data fit. Since these extended degrees of freedom are
not part of the basic physics chosen to model the data acquisition
process, they should be suppressed in the eventual solution. Extended
inversion methods differ by the choice of additional degrees of
freedom, and by choice of penalty applied to eliminate them in the final
result.

Many of these extension concepts sound
plausible, and appear to work at least to some extent as one might
hope from their heuristic justifications. However very few of these
approaches have been underwritten by mathematical argument: in
essence, they are mostly justified only ``in the rear-view mirror'',
with no assurance that failure is not just around the corner, at the
next example. On top of that, some of these approaches, for example
those based on the computationally attractive Variable Projection
Method (``VPM'') of \cite{GolubPereyra:03}, are cast in such form that the reasons for success are
not readily apparent.

This note shows exactly how VPM leads to successful velocity updates
for a particular extended inversion approach to a very simple inverse
problem, which asks that a homogeneous velocity field be deduced from
one trace at known offset. I put forward this inverse problem and
extension-based solution not because there are not simpler
ways of answering the question it poses - there certainly are - but
because the formal ingredients of waveform-based velocity estimation
in this very simple setting are common to many similar extended
inversion algorithms, and because in this case every computation can
be done analytically, nearly to completion. In particular, it becomes
clear why the VPM gradient formula produces a constructive update,
with no possibility of stagnation at local minimizers.

The extended inversion approach
developed here uses a {\em source extension}, in which source
parameters form the additional degrees of freedom. This type of
extension presumes that the actual or target source is constrained in
some way; the extended source is allowed to violate the
constraint. For the problem considered here, the source model reduces
to a wavelet, and the target wavelet is assumed to be non-zero only in
a short time
interval (an approximate impulse). The extension consists in permitting energy to spread
in time at intermediate iterations of the inversion. A simple penalty
for energy spread (second moment of square amplitude) drives the
extended source towards a focused source satisfying the assumed
constraint. Not all penalties are created equal: the penalty used here
has a critical {\em (pseudo-)differential} attribute necessary for avoidance of cycle-skipping.

@@

Ignoring amplitude, this map implements a $m$-dependent time
shift. This time shift operator is the basis of many descriptions of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above, which
amounts essentially to a time shift. To make the link with wavelet
frequency content manifest, I introduce a family $\{\fl\}$ of
wavelets parametrized by a nondimensional wavelength $\lambda$,
\begin{equation}
  \label{eqn:deffl}
  \fl(t) = \frac{1}{\lambda}f_1\left(\frac{t}{\lambda}\right).
\end{equation}
The ``mother wavelet'' $f_1$ vanishes for $t>1$, and has positive
mean-square. No other constraints need be placed on $f_1$. Note that
the scaling is such that the mean-square
\[
  \|\fl\|^2 = \int\,dt\,|\fl(t)|^2
\]
is independent of $\lambda$.

Note that in relations such as ``$t>1$'' a time unit is implicit in
the right-hand side.

@@

\section{Full Waveform Inversion}
The preceding section provided all of the raw ingredients to define
full waveform inversion for estimation of $m$ from a single trace.
It is only $m$ that is to be determined: the $\lambda-$dependent family of wavelets
$\{\fl\}$ is regarded as known, along with the data family $\{\dl\}$. The aim
is to chose $m$ to minimize 
\begin{equation}
  \label{eqn:FWIfix}
  J_{\rm fix}[m] = \frac{1}{2}\|S[m]\fl-\dl\|^2.
\end{equation}
for all values of $\lambda > 0$.

Written out in detail, this objective function is
\[
 J_{\rm fix}[m] =  \frac{1}{32\pi^2
    r^2}\int\,dt\,\left|\fl\left(t-mr\right)-\fl\left(t-m_*r\right)\right|^2
\]
Since $f_1$ vanishes for $|t|>1$, $\fl$ vanishes for $|t|>\lambda$,
and $S[m]\fl$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $S[m]\fl(t)=0$ or $S[m_*]\fl(t)=0$. Therefore $S[m]\fl$ and
$S[m_*]\fl$ are orthogonal in the sense of the $L^2$ inner product:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle S[m]\fl,
  S[m_*]\fl\rangle = \int\,dt\,S[m]\fl(t)S[m_*]\fl(t) = 0
\end{equation}
But $\dl = S[m_*]\fl$, so this is the same as saying that $\dl$ is
orthogonal to $S[m]\fl$. So conclude that
\begin{equation}
  \label{eqn:iso}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, J_{\rm fix}[m]=\frac{1}{16\pi^2
    r^2}\|f_1\|^2.
\end{equation}
using the previously observed independence of $\|\fl\|$ from
$\lambda$.

That is, for slowness $m$ in error by more than $\lambda/r$ from the
target slowness $m_*$, the FWI objective \ref{eqn:FWIfix} is perfectly
flat: all nearby values of $m$ are local minima. Therefore the local
exploration of the objective
gives no useful information whatever about constructive search
directions towards the global minimizer $m=m_*$, for which the
objective value is of course = 0, if the initial estimate of $m$ in
error by an amount greater than a fixed multiple of $\lambda$. This is
precisely the behaviour of FWI noted many times in the literature: the
model must be known to ``within a wavelength''.

\section{Source Extension Inversion}
An obvious extended model for this example is $\{m,f\}$, that is, pairs of velocity and source wavelet, with the extended modeling operator already defined by equation \ref{eqn:mod}. With an annihilator $A$, to be chosen below, the penalty form of extended inversion is: minimize over $\{v,f\}$
\begin{equation}
\label{eqn:obj}
J[v,f] = \frac{1}{2}(\|S[m]f-d\|^2 + \alpha^2 \|Af\|^2).
\end{equation}
VPM \cite[]{GolubPereyra:03} takes advantage of $J$ being quadratic in $f$ to solve for $f$ given $v$, thus producing a reducted objective of $v$ alone:
\begin{equation}
\label{eqn:red}
J_{\rm red}[v] = \min_f J[v,f] = J[v,f[v]],
\end{equation}
where $f[v]$ is the minimizer of $J$ over $f$ for given $v$.
For this example, $J_{\rm red}$ is simple to compute. First observe that apart from amplitude, $S[m]$ is unitary:
\begin{equation}
\label{eqn:tran}
S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+\frac{r}{v}\right)
\end{equation}
so
\begin{equation}
\label{eqn:unit}
S[m]^TS[m] f (t) = \frac{1}{(4\pi r)^2} f(t).
\end{equation}
Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:red} is
\begin{equation}
\label{eqn:norm1}
\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f = S[m]^Td.
\end{equation}

At this point I have to come clean about the actual choice of $A$. As usual with source extensions, the proper annihilator depends on modeling assumptions, not fundamental physics. Assume that signature decon has been applied and the target source $f_*$ is a zero-phase bandpass filter, reasonably well focused at time $t=0$. Since there is no spatial structure to exploit in this hyper-simple problem, we construct $A$ to penalize energy away from $t=0$:
\begin{equation}
\label{eqn:ann}
Af(t) = tf(t).
\end{equation}
Then the normal operator on the LHS of \ref{eqn:norm1} is simply multiplication by a positive function of time, and can be inverted by inspection. Using the identity \ref{eqn:tran} for the adjoint operator, obtain
\begin{equation}
\label{eqn:soln} 
f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{4\pi r}d\left(t+\frac{r}{v}\right)  
\end{equation} 
Suppose that the data is noise-free: 
\begin{equation}
\label{eqn:nonoise}
d(t) = S[v_*]f_*(t) = \frac{1}{4\pi r}f_*\left(t-\frac{r}{v_*}\right)
\end{equation}
Then
\begin{equation}
\label{eqn:solnnon}
f[v](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right) 
\end{equation}
so the residual is, after a little algebra,
\begin{equation}
\label{eqn:resid}
(S[m]f[v]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-r/v)^2)^{-1} - 1]f_*(t-r/v_*)
\end{equation}
The gradient of $J_{\rm red}$ is given by the famous VPM formula (well-known to those who know it well):
\begin{equation}
\label{eqn:grad}
\nabla J[v] = (DS[m]f[v])^*(S[m]f[v]-d)
\end{equation}.
In this formula, $DS[m]$ is the derivative with respect to $v$ - since we can make explicit use of the Green's function, easy to see that
\begin{equation}
\label{eqn:deriv}
(DS[m]\delta v)f (t) = \frac{\delta v}{4\pi v^2}\frac{df}{dt}(t-r/v) = S[m](Q[v]\delta v)f (t),
\end{equation}
where 
\begin{equation}
\label{eqn:defq}
(Q[v]\delta v)f = \frac{r \delta v}{v^2} \frac{df}{dt}.
\end{equation}
That is, $Q[v]\delta v$ is a skew-adjoint operator depending linearly on $\delta v$ - more on this below.
The adjoint $(DS[m]f)^*$ is the adjoint of the map from model space to data space:
\[
(DS[m]f): \delta v \mapsto (DS[m]\delta v)f
\]
and is NOT the same as the adjoint denoted by $S[m]^T$, which is the data-space-to-data-space adjoint (got that?).
That is, this adjoint make this relation true, for all $\delta v,f,$ and $d$:
\begin{equation}
\label{eqn:vadj}
\delta v \cdot (DS[m]f)^*d = \int \,dt\,[(DS[m]\delta v)f] (t) d(t) 
\end{equation}
On the left side is the dot product in model space - since model space is just 1D in this example, that's just the numerical product. On the right is the dot product in data space, in the idealized continuum limit. Using equations \ref{eqn:deriv}, \ref{eqn:defq} and \ref{eqn:vadj},
\[
\delta v \cdot \nabla J[v] =  \int \,dt\,[(DS[m]\delta v)f[v]] (t) (S[m]f[v]-d)(t)
\]
\[
= \int \,dt\, [S[m](Q[v]\delta v)f[v](t)] (S[m]f[v]-d)(t) = \int
\,dt\,(Q[v]\delta v)f[v](t) S[m]^T(S[m]  f[v]-d)(t)
\]
Now invoke the normal equation \ref{eqn:norm1}: and replace the last factor:
\begin{equation}
\label{eqn:comm1}
= -\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t)
\end{equation}
Since $Q[v]\delta v$ is skew-symmetric, shift it onto the other factor in this $L^2$ inner product (why not):
\[
= \alpha^2\int \,dt\, f[v](t)(Q[v]\delta v)A^TAf[v](t) 
\]
\begin{equation}
\label{eqn:comm2}
= \alpha^2\int \,dt\, (A^TAf[v](t)(Q[v]\delta v)f[v](t) + f[v](t)[(Q[v]\delta v),A^TA]f[v](t))
\end{equation}
where I swapped $Q$ and $A^TA$ at the cost of introducing a term involving the commutator $[Q,A^TA] = QA^TA-A^TAQ$, and rearranged the first term using the symmetry of $A^TA$. Now notice that this first term is exactly the same as the RHS of equation \ref{eqn:comm1}, except for the minus sign - so subtracting the RHS of \ref{eqn:comm2}
from \ref{eqn:comm1} and rearranging, get
\[
-\alpha^2\int \,dt\, (Q[v]\delta v)f[v](t)A^TAf[v](t) = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
\]
hence 
\begin{equation}
\label{eqn:gradcomm}
\delta v \cdot \nabla J[v] = \frac{1}{2}\alpha^2\int \,dt\,f[v](t)[(Q[v]\delta v),A^TA]f[v](t)
\end{equation}
Remember that $A^TA$ amounts to multiplying by $t^2$, and $Q$ is the scaled time derivative (equation \ref{eqn:defq}), so
\begin{equation}
\label{eqn:comm3}
[(Q[v]\delta v),A^TA]=\frac{2r\delta v}{v^2}t
\end{equation}
Insert this identity into equation \ref{eqn:gradcomm} to obtain
\begin{equation}
\label{eqn:comm4}
\delta v \cdot \nabla J_{\rm red}[v] =  \delta v\frac{r \alpha^2}{v^2}\int \,dt\,tf[v]^2(t)
\end{equation}
Combine this identity with the formula \ref{eqn:solnnon} for the solution of the inner problem, and divide out the common factor $\delta v$, to obtain
\[
\nabla J_{\rm red}[v] = \frac{r\alpha^2}{v^2}\int \,dt\,t\left(\frac{1}{(4\pi r)^2} + \alpha^2 t^2\right)^{-1}\frac{1}{(4\pi r)^2}f_*\left(t-\frac{r}{v_*}+\frac{r}{v}\right)^2
\]
\begin{equation}
\label{eqn:gradfinal}
= \frac{r\alpha^2}{v^2}\int \,dt\,\left(t+\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(t+\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}f_*(t)^2
\end{equation}

Recall that the definition of $A$ (multiplication by $t$) was justified by a modeling assumption, that the actual source is (possibly after designature) highly localized at $t=0$ - ideally a Dirac delta, but because of bandlimitation not quite. Still, a high degree of localization, as implicit in this choice of annihilator, means that the integral above is approximately proportional to the integrand at $t=0$, that is, 
\[
\approx \frac{r\alpha^2}{v^2}\left(\frac{r}{v_*}-\frac{r}{v}\right)\left(1 + (4\pi r)^2\alpha^2 \left(\frac{r}{v_*}-\frac{r}{v}\right)^2\right)^{-1}
\]
Note that to the extent that this approximation is accurate,
\begin{itemize}
\item if $v > v_*$, then $\nabla J_{\rm red}[v] > 0$, and
\item if $v < v_*$, then $\nabla J_{\rm red}[v] < 0$, and
\end{itemize}
That is, within the domain of validity of this approximation (setting $t=0$ in equation \ref{eqn:gradfinal}), {\em there are no local mins:} the gradient has the correct sign and velocity updates computed from it will be constructive.

On the other hand, careful examination of equation \ref{eqn:gradfinal} (which involves no approximations) shows that if $\alpha$ is too large, then the finite-frequency spread of $f_*$ is guaranteed to result in non-convex behaviour, since the side-lobes will dominate. This message is consistent with the import of many other similar computations.

\section{Discussion}
There remain two important points to be made.

First, the formal computations centering on the operator $Q$ (equations \ref{eqn:deriv} through \ref{eqn:gradcomm}) depend only on the relation \ref{eqn:deriv} and the skew-symmetry of $Q$, which hold with minor modifications for other more complex waveform inversion problems. These other more complex problems do not submit to such a simple treatment as is shown in the equations following \ref{eqn:gradcomm}, but for example it is possible in some cases to use the relation \ref{eqn:deriv} to extract a relation between extended waveform inversion and traveltime tomography, via analysis of the Hessian at a zero-residual global minimizer. See for instance \cite{tenKroode:IPTA14,Symes:IPTA14,Symes:Madrid}. Up to that point the reasoning is quite general, and central to the understanding gained so far of extended inversion methods.

>>>>

\maketitle
\begin{abstract}
  An extremely simple single-trace transmission example shows how an
  extended source formulation of full waveform inversion can produce
  an optimization problem without spurious local minima (``cycle
  skipping''). The data consist of a single trace recorded at a given
  distance from a point source. The velocity or slowness is presumed
  homogeneous, and the target source wavelet is presumed
  quasi-impulsive or focused at zero time lag. The source is extended
  by permitting energy to spread in time, and the spread is controlled
  by adding a weighted mean square of the extended source wavelet to
  the data misfit, to produce the extended inversion objective. The
  objective function and its gradient can be computed explicitly, and
  it is easily seen that all local minimizers must be within a
  wavelength of the correct slowness. The derivation shows several
  important features of all similar extended source algorithms. For
  example, nested optimization, with the source estimation in the
  inner optimization (variable projection method), is essential. The
  choice of the weight operator, controlling the extended source
  degrees of freedom, is critical: the choice presented here is a
  differential operator, and that property is crucial for production
  of an objective immune from cycle-skipping.
\end{abstract}

\section{Introduction}
Full Waveform Inversion (FWI), or estimation of earth structure by
model-driven least squares data fitting, is now well-established as a
useful tool for probing the earth's subsurface
\cite[]{VirieuxOperto:09,Fichtner:10}. However, so-called ``cycle-skipping'', the tendency of iterative FWI
algorithms to stagnate at suboptimal and geologically uninformative
earth models, still impedes its use. Because the computational size of field inversion tasks
is very large, only iterative local (descent) minimization of the data
misfit function is computationally feasible. However local
descent methods avoid suboptimal stagnation only if initial models are
already quite close to optimal, in the sense of predicting the arrival
times of seismic events to within a small multiple of a dominant
wavelength \cite[]{GauTarVir:86,Plessix:10}.

This paper concerns one of the several ideas that have been advanced to
overcome cycle-skipping, namely so-called extended inversion
\cite[]{geoprosp:2008}. ``Extended'' signifies that addional degrees
of freedom are provided to the modeling process, in the hope of
opening up more effective routes to geologically informative models
with acceptable data fit. Since these extended degrees of freedom are
not part of the basic physics chosen to model the data acquisition
process, they should be suppressed in the eventual solution. Extended
inversion methods differ by the choice of additional degrees of
freedom, and by choice of penalty applied to eliminate them in the final
result.

Many of these extended inversion concepts sound plausible, and appear
to work at least to some extent as one might hope from their heuristic
justifications. However very few of these approaches have been
underwritten by mathematical argument: in essence, they are mostly
justified only ``in the rear-view mirror'', with no assurance that
failure is not just around the corner, at the next example. On top of
that, some of these approaches, for example those based on the
computationally attractive Variable Projection Method (``VPM'') of
\cite{GolubPereyra:03}, are cast in such form that the reasons for
success are not readily apparent.

This note shows exactly how VPM leads to successful velocity updates
for a particular extended inversion approach to a very simple inverse
problem, which asks that a homogeneous velocity field be deduced from
one trace at known offset. I put forward this inverse problem and
extension-based solution not because there are not simpler
ways of answering the question it poses - there certainly are - but
because the formal ingredients of waveform-based velocity estimation
in this very simple setting are common to many similar extended
inversion algorithms, and because in this case every computation can
be done analytically, nearly to completion. In particular, it becomes
clear why the VPM gradient formula produces a constructive update,
with no possibility of stagnation away from the global minimum.

The extended inversion approach
developed here uses a {\em source extension}, in which source
parameters form the additional degrees of freedom. This type of
extension presumes that the actual or target source is constrained in
some way; the extended source is allowed to violate the
constraint. For recent overview of source extension methods, see
\cite{HuangNammourSymesDollizal:SEG19}. Source
extension methods have computational complexity approximately the same
as that of FWI, a signal advantage over the alterantive medium
extension class described for instance in \cite[]{geoprosp:2008}.

For the problem considered here, the source model amounts 
to a wavelet, and the target wavelet is assumed to be non-zero only in
a short time
interval (an approximate impulse, perhaps as the result of signature deconvolution). The extension consists in permitting energy to spread
in time at intermediate iterations of the inversion. A simple penalty
for energy spread (second moment of square amplitude) drives the
extended source towards a focused source approximately satisfying the assumed
constraint. Not all penalties are created equal: the penalty used here
has the critical {\em (pseudo-)differential} attribute necessary for
avoidance of cycle-skipping, as will be explained in the Discussion section.

@@

Ignoring amplitude, this map implements a $m$-dependent time
shift. This time shift operator is the basis of many descriptions of
the cycle-skipping phenomenon (for example, \cite{VirieuxOperto:09},
Figure 7), so it is unsurprising that an analysis of cycle-skipping
can be based on the simple modeling operator described above, which
amounts essentially to a time shift. To make the link with wavelet
frequency content manifest, I introduce a family $\{\fl\}$ of
wavelets indexed by $\lambda$, a parameter having dimensions of time,
\begin{equation}
  \label{eqn:deffl}
  \fl(t) = \frac{1}{\lambda}f_1\left(\frac{t}{\lambda}\right).
\end{equation}
The argument $s$ of the ``mother wavelet'' $f_1$ is
nondimensional. The only constraints placed on $f_1$ are that (i) 
$f_1(s)=0$ for $|s|\ge 1$, and (ii) $f_1$  has positive
mean-square, that is, does not vanish identically. Note that
the scaling is such that the mean-square
\[
  \|\fl\|^2 = \int\,dt\,|\fl(t)|^2
\]
is independent of $\lambda$.

I shall refer to $\lambda$ as ``wavelength'': if $f_1$ has a dominant
period of oscillation, then so does $\fl$, and it is proportional to
$\lambda$.

@@

\section{Full Waveform Inversion}
The preceding section provided all of the raw ingredients to define
full waveform inversion for estimation of $m$ from a single trace.
It is only $m$ that is to be determined: the $\lambda-$dependent family of wavelets
$\{\fl\}$ is regarded as known, along with the data family $\{\dl\}$. The aim
is to chose $m$ to minimize 
\begin{equation}
  \label{eqn:FWIfix}
  J_{\rm FWI}[m] = \frac{1}{2}\|S[m]\fl-\dl\|^2.
\end{equation}
for all values of $\lambda > 0$.

Written out in detail, this objective function is
\[
 J_{\rm FWI}[m] =  \frac{1}{32\pi^2
    r^2}\int\,dt\,\left|\fl\left(t-mr\right)-\fl\left(t-m_*r\right)\right|^2
\]
Since $f_1$ vanishes for $|t|>1$, $\fl$ vanishes for $|t|>\lambda$,
and $S[m]\fl$ vanishes if $|t-mr|>\lambda$. So if $|mr-m_*r|
= |m-m_*|r > 2\lambda$, then $|t-mr|+|t-m_*r| \ge |mr-m_*r| >
2\lambda$ so either $|t-mr|>\lambda$ or $|t-m_*r|>\lambda$, that is,
either $S[m]\fl(t)=0$ or $S[m_*]\fl(t)=0$. Therefore $S[m]\fl$ and
$S[m_*]\fl$ are orthogonal in the sense of the $L^2$ inner product:
\begin{equation}
  \label{eqn:ortho}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, \langle S[m]\fl,
  S[m_*]\fl\rangle = \int\,dt\,S[m]\fl(t)S[m_*]\fl(t) = 0
\end{equation}
But $\dl = S[m_*]\fl$, so this is the same as saying that $\dl$ is
orthogonal to $S[m]\fl$. So conclude that
\begin{equation}
  \label{eqn:iso}
  |m- m_*|r > 2\lambda \,\, \Rightarrow \,\, J_{\rm FWI}[m]=\frac{1}{16\pi^2
    r^2}\|f_1\|^2.
\end{equation}
using the previously observed independence of $\|\fl\|$ from
$\lambda$.

That is, for slowness $m$ in error by more than $2\lambda/r$ from the
target slowness $m_*$, the FWI objective $J_{\rm FWI}$ is perfectly
flat: all nearby values of $m$ are local minima. Therefore the local
exploration of the objective
gives no useful information whatever about constructive search
directions towards the global minimizer $m=m_*$, for which the
objective value is of course = 0, if the initial estimate of $m$ in
error by an amount greater than a fixed multiple of $\lambda$. This is
precisely the behaviour of FWI noted many times in the literature: the
model must be known to ``within a wavelength''.

\section{Extended Source Inversion}
As mentioned in the introduction, the modeling operator introduced in
the last section, $m \mapsto S[m]\fl$, may be extended simply by
including the source wavelet as one of the model parameters: that is,
the model vector becomes $(m,f)$, and the modeling operator, $(m,f)
\mapsto S[m]f$.

The reader will have no trouble seeing that the data misfit using this
extended modeling operator can always be made to vanish entirely by
proper choice of wavelet $f$, unless $f$ must satisfy some additional
constraints. \cite{HuangNammourSymesDollizal:SEG19} describe a
plethora of possible constraints for this and similar source
extensions. Many (but not all) take the form of a quadratic penalty,
that is, the mean square of $Af$, $A$ being a suitable operator,
commonly dubbed an {\em annihilator}: in many examples the ideal
output for $A$ applied to a source obeying the target constraints
is the zero vector \cite[]{geoprosp:2008}. With an annihilator $A$, to
be chosen below, the penalty form of extended inversion is: minimize
over $\{m,f\}$
\begin{equation}
\label{eqn:obj}
J_{\rm ESI}[m,f] = \frac{1}{2}(\|S[m]f-\dl\|^2 + \alpha^2 \|Af\|^2).
\end{equation}

The choice of penalty weight $\alpha$ has a profound influence on the
character of this optimization problem. For the moment, I will mandate
only that $\alpha > 0$. A detailed discussion of methods for setting
$\alpha$ is beyond the scope of this brief paper.  In the Discussion
section, I mention an effective method for choosing $\alpha$ to
optimize convergence of iterative minimization, that applies (and has
been applied) to much larger-scale inversion problems with features
similar to those of this simple example.

While minimization of $J_{\rm ESI}$ might be tackled directly - by
alternately minimizations between $m$ and $f$, or by computing updates
for $m$ and $f$ simultaneously - such joint mimization performs
poorly, as \cite{YinHuang:16} has shown. The reason for this poor
performance is that $J_{\rm ESI}$ has dramatically different
sensitivity to $m$ versus $f$, especially for high frequency $f$, as
the reader will see below. Instead, a nested approach, in which $f$ is
eliminated in an inner optimization,
generally gives far better numerical performance.  This Variable
Projection Method (VPM) \cite[]{GolubPereyra:03} takes advantage of
$J_{\rm ESI}$ being quadratic in
$f$ to solve for $f$ given $m$, thus producing a reduced objective of
$m$ alone:
\begin{equation}
\label{eqn:red}
J_{\rm VPM}[m] = \min_f J_{\rm ESI}[m,f] = J_{\rm ESI}[m,f[m]],
\end{equation}
where $f[m]$ is the minimizer of $J_{\rm ESI}$ over $f$ for given $m$.
For the problem considered here, $J_{\rm VPM}$ is explicitly computable. First observe that apart from amplitude, $S[m]$ is unitary:
\begin{equation}
\label{eqn:tran}
S[m]^T g (t) = \frac{1}{4\pi r}g\left(t+mr\right)
\end{equation}
so
\begin{equation}
\label{eqn:unit}
S[m]^TS[m] f (t) = \frac{1}{(4\pi r)^2} f(t).
\end{equation}
Therefore the normal equation for the minimizer on the RHS of equation \ref{eqn:red} is
\begin{equation}
\label{eqn:norm1}
\left(\frac{1}{(4\pi r)^2} I + \alpha^2 A^TA\right)f[m] = S[m]^T\dl.
\end{equation}

At this point I have to come clean about the actual choice of
$A$. Recall that $A$ is called an ``annihilator'', as it vanishes when
applied to the target source field, at least in an idealized
limit. Thus the choice of $A$ depends on modeling assumptions, not
fundamental physics, as is characteristic of extended source
inversion. As $\lambda \rightarrow 0$, the target source wavelet $\fl$
focuses at time $t=0$, in the sense that it vanishes for
$|t| > \lambda$. Therefore I choose $A$ to penalize energy away from
$t=0$:
\begin{equation}
\label{eqn:ann}
Af(t) = tf(t).
\end{equation}
This
particular annihilator has been employed in earlier papers on extended
source inversion
\cite[]{Plessix:00a,LuoSava:11,Warner:14,HuangSymes:SEG15a,Warner:16,HuangSymes:GEO17}.

With this choice of $A$, the normal operator on the LHS of \ref{eqn:norm1} is simply multiplication by a positive function of time, and can be inverted by inspection. Using the identity \ref{eqn:tran} for the adjoint operator, obtain
\[
f[m](t) = \left(\frac{1}{(4\pi r)^2} + \alpha^2
  t^2\right)^{-1}\frac{1}{4\pi r}\dl\left(t+mr\right)
\]
\begin{equation}
\label{eqn:solnnon}
= \left(1+ (4\pi r)^2\alpha^2 t^2\right)^{-1}\fl\left(t+(m-m_*)r\right) 
\end{equation}
thanks to the definition of the data $\dl$ (equation \ref{eqn:defdata}).
After a little algebra,
\begin{equation}
\label{eqn:resid}
(S[m]f[m]-d)(t) =\frac{1}{4\pi r}[ (1+(4\pi r)^2 \alpha^2 (t-mr)^2)^{-1} - 1]\fl(t-m_*r)
\end{equation}
and
\begin{equation}
  \label{eqn:anniexp}
  Af[m] = t \left(1+ (4\pi r)^2\alpha^2
    t^2\right)^{-1}\fl\left(t+(m-m_*)r\right).
\end{equation}
So
\[
  J_{\rm VPM}[m] = \frac{1}{2}\int\,dt\, \left( \frac{1}{(4\pi r)^2}[
    (1+(4\pi r)^2 \alpha^2 (t+(m_*-m)r)^2)^{-1} - 1]^2 +
  \right.
  \]
  \[
   \left. \alpha^2(t+(m_*-m)r)^2 \left(1+ (4\pi r)^2\alpha^2
      (t+(m_*-m)r)^2\right)^{-2}\right)|\fl(t)|^2
\]
\begin{equation}
\label{eqn:vpmexpl}
=\frac{1}{2(4\pi r)^2}\int\,dt\, [1-(1+(4\pi r)^2 \alpha^2
(t+(m_*-m)r)^2)^{-1}]|\fl(t)|^2
\end{equation}

The gradient of $J_{\rm VPM}$ can be extracted by elementary means
from the identity \ref{eqn:vpmexpl}, but instead we will derive it in
using some important features that this inverse problem shares with
other inverse wave problems with more complex physics. To begin with, the gradient of a VPM
objective of the form \ref{eqn:red} is given by the formula 
\begin{equation}
\label{eqn:grad}
\nabla J_{\rm VPM}[m] = (DS[m]f[m])^*(S[m]f[m]-d).
\end{equation}
This easily derived result is in some sense the main content of \cite[]{GolubPereyra:03}.  In this formula, $DS[m]f$ is the derivative
of the modeling operator $S[m]f$
with respect to $m$, that is, with $f$ held fixed. The adjoint $(DS[m]f)^*$ is the adjoint of the map from model space to data space:
\[
(DS[m]f): \delta m \mapsto (DS[m]\delta m)f
\]
and is NOT the same as the adjoint denoted by $S[m]^T$, which is the source-space-to-data-space adjoint.
That is, this adjoint make this relation true, for all $m, \delta m,f,$ and $d$:
\begin{equation}
\label{eqn:vadj}
\delta m \cdot (DS[m]f)^*d = \int \,dt\,[(DS[m]\delta m)f] (t) d(t) 
\end{equation}
On the left side is the dot product in model space - since model space
is just 1D in this example, that's just the numerical product. On the
right is the dot product in data space, in the idealized continuum
limit.

Note that the VPM gradient formula is remarkable in two ways. First,
it is exactly the same as the FWI gradient, that is, the gradient of
$J_{\rm FWI}$ if you happen to insert the solution $f[m]$ of the
normal equation for $f$. Second, it does not mention the annihilator
$A$ at all: its impact is locked up in $f[m]$.

The key to unlocking the meaning of the VPM gradient formula for this
and similar problems is a remarkable feature of the derivative
operator $DS[m]$: from the definition \ref{eqn:mod},
\begin{equation}
\label{eqn:deriv}
(DS[m]\delta m)f (t) = -\frac{\delta m}{4\pi}\frac{df}{dt}(t-mr) = S[m](Q[m]\delta m)f (t),
\end{equation}
where 
\begin{equation}
\label{eqn:defq}
(Q[m]\delta m)f = -r\delta m \frac{df}{dt}.
\end{equation}
That is, $Q[m]\delta m$ is a skew-adjoint operator depending linearly
on $\delta m$ - more on this below.

A calculation, detailed in Appendix A, yields an expression for the
gradient \ref{eqn:grad} in terms of $Q$:
\begin{equation}
\label{eqn:gradcomm}
\delta m \cdot \nabla J_{\rm VPM}[m] = \frac{1}{2}\alpha^2\int \,dt\,f[m](t)[(Q[m]\delta m),A^TA]f[m](t)
\end{equation}
Here, the symbol $[L,M]$ denotes the {\em commutator} of the operators
$L$ and $M$: $[L,M]=LM-ML$.

Note that the annihilator $A$ is explicitly present in
\ref{eqn:gradcomm}. The structure displayed in this expression is
common to many other extended inversion methods. The extreme
simplicity of the factorization \ref{eqn:deriv}, \ref{eqn:defq} and
the gradient expression \ref{eqn:gradcomm} is
modified for more complex inversion problems by asymptotically
negligible corrections \cite[]{Symes:IPTA14,tenKroode:IPTA14,Symes:Madrid}.

%%%%%%%%%%%
